sampbias, a method for quantifying geographic sampling biases in species distribution data

### 2020-07-31 04_appendix_S1.Rmd
manuscript updates and script re-organization

## get a polygon of Borneo
data(landmass)

born2 <- st_read("empirical_analyses/Borneo.kml", quiet = TRUE)
born2 <- sf:::st_zm(born2$geom)
born2 <- as(born2, 'Spatial')

@@

# prepare for plotting
airp <- airports %>% crop(SpatialPoints(occ[,2:1])) %>% raster::intersect(born2) %>%  coordinates() %>%  data.frame() %>% mutate(type = "Airport")
cit <- cities %>% crop(SpatialPoints(occ[,2:1])) %>% raster::intersect(born2) %>%  coordinates() %>%  data.frame() %>%  mutate(type = "City")
pts <- bind_rows(occ, cit, airp)

riv <- waterbodies %>% crop(SpatialPoints(occ[,2:1])) %>%  fortify() %>%  mutate(type = "River")
road <- roads %>% crop(SpatialPoints(occ[,2:1])) %>% raster::intersect(born2) %>%  fortify() %>%  mutate(type = "Road")
lin <- bind_rows(riv, road)

lma <- landmass %>% crop(SpatialPoints(occ[,2:1])) %>% raster::intersect(born2) %>%  fortify()

@@

```{r}
occ <-read.csv(system.file("extdata", "mammals_borneo.csv", package="sampbias"), sep = "\t")

data(landmass)

born <- raster::intersect(landmass, born2)

area <- round(areaPolygon(born) / 1000000, 0)

REWRITE THIS PART TO DEAL WITH THE NEW TYPE OF OUTPUT FILES

dat <- read_csv("empirical_analyses/simulations/weight_estimates.csv") %>% 
  mutate(records_area = round(rar / area * 10000, 2)) %>% 
  select(contains("w"), ID, type, records_area) %>% 
  select(-w_rate) %>% 
    pivot_longer(cols = contains("w_"), names_to = "bias",
                 values_to = "posterior_estimate") %>%
    mutate(bias = gsub("w_", "", .data$bias))
```

```{r, echo = FALSE, fig.cap="The bias weights ($w$) quantifying the effects of each bias factor estimated from data sets with differing density of occurrence records across the study area. Datasets generated by down-sampling the empirical example dataset of mammal occurrences on Borneo (www.gbif.org,  2016, https://doi.org/10.15468/dl.7fg4zx)."}
plo <- dat %>%  filter(type == "empirical") %>%
    mutate(bias = fct_reorder(.data$bias, .data$posterior_estimate, .fun = median, .desc = FALSE))
# ggplot(data = plo)+
#   geom_point(aes(x = bias, 
#                  y = posterior_estimate, 
#                  group = as.factor(records_area), 
#                  color = as.factor(ID)), 
#              position = position_jitterdodge())+
#   xlab("Posterior weight")+
#   xlab("Bias factor")+
#   scale_color_viridis(discrete = TRUE)+
#   theme_bw()

ggplot(data = plo)+
  geom_boxplot(aes(x = bias, 
                   y = posterior_estimate,
                 fill = as.factor(records_area),
                 color = as.factor(records_area)))+
  ylab("Posterior weight")+
  xlab("Bias factor")+
  scale_fill_viridis(discrete = TRUE, name = "Records per\n10,000 sqkm")+
  scale_color_viridis(discrete = TRUE, name = "Records per\n10,000 sqkm")+
  theme_bw()
```

@@

REDO WITH THE NEW RESULTS

@@

```{r, echo = FALSE, fig.cap="The bias weights ($w$) defining the effects of each bias factor estimated from data sets with differing density of occurrence records across the study area. Datasets generated by random simulation of records across Borneo (no bias)."}

Here a boxplot comparing the results form the empirical analyses with the ones fromt he randomly simulated analyses

plo <- dat %>%  filter(type == "simulated") %>%
    mutate(bias = fct_reorder(.data$bias, .data$posterior_estimate, .fun = median, .desc = FALSE))

ggplot(data = plo)+
  geom_boxplot(aes(x = bias, 
                   y = posterior_estimate,
                   fill = as.factor(records_area),
                   color = as.factor(records_area)))+
  ylab("Posterior weight")+
  xlab("Bias factor")+
  scale_fill_viridis(discrete = TRUE, name = "Records per\n10,000 sqkm")+
  scale_color_viridis(discrete = TRUE, name = "Records per\n10,000 sqkm")+
  theme_bw()
```

```{r, echo = FALSE, fig.height = 8, fig.cap="The bias weights ($w$) defining the effects of each bias factor estimated for the smallest simulated dataset without bias."}

Here a histogram of the weight estimates

ggplot(data = plo)+
  geom_histogram(aes(posterior_estimate, fill = as.factor(records_area)), bins = 50) +
  scale_fill_viridis(discrete = TRUE, name = "Records per\n10,000 sqkm")+
  facet_grid(rows = vars(records_area), cols = vars(bias))+
  scale_x_continuous(breaks = c(0,0.003, 0.006, 0.009), limits = c(0,0.01))+
  xlab("Posterior weight")+
  ylab("Count")+
  theme_bw()+
  theme(panel.grid.minor = element_blank(),
        legend.position = "none")

>>>>

## get a polygon of Borneo
data(borneo)

@@

# prepare for plotting
airp <- airports %>% crop(SpatialPoints(occ[,2:1])) %>% raster::intersect(borneo) %>%  coordinates() %>%  data.frame() %>% mutate(type = "Airport")
cit <- cities %>% crop(SpatialPoints(occ[,2:1])) %>% raster::intersect(borneo) %>%  coordinates() %>%  data.frame() %>%  mutate(type = "City")
pts <- bind_rows(occ, cit, airp)

riv <- waterbodies %>% crop(SpatialPoints(occ[,2:1])) %>%  fortify() %>%  mutate(type = "River")
road <- roads %>% crop(SpatialPoints(occ[,2:1])) %>% raster::intersect(borneo) %>%  fortify() %>%  mutate(type = "Road")
lin <- bind_rows(riv, road)

lma <- borneo %>%  fortify()

@@

```{r}
occ <-read.csv(system.file("extdata", "mammals_borneo.csv", package="sampbias"), sep = "\t")
area <- round(areaPolygon(borneo) / 1000000, 0)
```

```{r, echo=FALSE}

inp <- paste("empirical_analyses/simulations/", list.files("empirical_analyses/simulations/", pattern = "weightestimate"), sep = "")
inp <- lapply(inp, "read_csv")

dat <- bind_rows(inp) %>% 
  mutate(records_area = round(rar / area * 10000, 2)) %>% 
  dplyr::select(contains("w"), ID, type, records_area) %>%
  pivot_longer(cols = contains("w_"), names_to = "bias",
                 values_to = "posterior_estimate") %>%
    mutate(bias = gsub("w_", "", .data$bias)) %>% 
    mutate(bias = fct_reorder(.data$bias, .data$posterior_estimate, .fun = median, .desc = FALSE))

```{r, echo = FALSE, fig.height = 7, fig.cap="The bias weights ($w$) of the empirical analyses of records from mammals from Borneo (6262 records, www.gbif.org, 2016, https://doi.org/10.15468/dl.7fg4zx) in comparison to ten simulated datasets of the same size with randomly picked locations across Borneo (no bias)."}

emp <- dat %>% 
  filter(records_area == 85) %>% 
  mutate(ID = "Empirical") %>% 
  mutate(type = "Empirical") %>%
  dplyr::select(-records_area)

sim <- paste("empirical_analyses/simulations/", list.files("empirical_analyses/simulations/", pattern = "biasweights"), sep = "")
sim <- lapply(sim, "read_csv")

plo <- bind_rows(sim) %>%
  mutate(ID = paste("Replicate_", as.character(ID), sep = "")) %>% 
  mutate(type = "Simulated (no bias)") %>% 
  dplyr::select(contains("w"), ID, type) %>%
  pivot_longer(cols = contains("w_"), names_to = "bias",
                 values_to = "posterior_estimate") %>%
    mutate(bias = gsub("w_", "", .data$bias)) %>% 
  bind_rows(emp)

ggplot(data = plo)+
  geom_boxplot(aes(x = ID, 
                   y = posterior_estimate,
                 fill = as.factor(type),
                 color = as.factor(type)))+
  ylab("Posterior weight")+
  scale_fill_viridis(discrete = TRUE)+
  scale_color_viridis(discrete = TRUE)+
  theme_bw()+
  theme(legend.title = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "bottom")+
  facet_wrap(.~bias, ncol = 1)
```

```{r, echo = FALSE, fig.cap="The bias weights ($w$) quantifying the effects of each bias factor estimated from data sets with differing density of occurrence records across the study area. Datasets generated by down-sampling the empirical example dataset of mammal occurrences on Borneo (www.gbif.org,  2016, https://doi.org/10.15468/dl.7fg4zx)."}
ggplot(data = dat)+
  geom_boxplot(aes(x = bias, 
                   y = posterior_estimate,
                 fill = as.factor(records_area),
                 color = as.factor(records_area)))+
  ylab("Posterior weight")+
  xlab("Bias factor")+
  scale_fill_viridis(discrete = TRUE, name = "Records per\n10,000 sqkm")+
  scale_color_viridis(discrete = TRUE, name = "Records per\n10,000 sqkm")+
  theme_bw()
```

@@

@@
